Designing Qusecs Optimally: Quasi-uniform or Self-similar Layered Materials as Recursively Decomposed Solutions of Geometric Constraint Systems

### 2015-05-17 doc/GMP2015/revision_2/s04_recombination.tex
Revision 2, finished.

\subsubsection{Optimal Recombination and Solution Space Navigation.}
For the entire DR-plan, finding all desired solutions is barely tractable even if recombination of solved subsystems is easy for each indecomposable parent system in the DR-plan. This is because even for the simplest, highly decomposable systems with small DR-plans, the problem of finding even a single solution to the input system at the root of the DR-plan is NP-hard  \cite{saxe1979embeddability} and there is a combinatorial explosion of solutions \cite{borcea2004number}. Typically, however, the desired solution has a given orientation type, in which case, the crux of the difficulty is concentrated in the algebraic complexity of (re)combining child system solutions to give a solution to the parent system at any given level of the DR-plan. For fairly general 3D constraint systems, there are algorithms with formal guarantees that uncover underlying matroids to combinatorially obtain an optimal parameterization to minimize the algebraic complexity of the indecomposable parent (sub)systems that occur in the DR-plan \cite{sitharam2010optimized,sitharam2006well,sitharam2010reconciling}, provided the DR-plan meets some of the abovementioned criteria.

@@

\subsubsection{Configuration Spaces of Underconstrained Systems.}
For underconstrained 2D bar-joint and body-hyperpin qusecs obtained from various subclasses of tree-decomposable systems, algorithms have been developed to complete them into isostatic systems
\cite{joan-arinyo2003transforming,sitharam2005combinatorial,gao2006ctree,sitharam2010convex} and to find paths within the connected components \cite{sitharam2011cayleyI,hidalgo2011reachability} of standard Cartesian configuration spaces. Most of the algorithms with formal guarantees leverage Cayley configuration space theory \cite{sitharam2010convex,sitharam2011cayleyI,sitharam2011cayleyII} to characterize subclasses of graphs and additional constraints that control combinatorial explosion, and provide faithful bijective representation of connected components and paths. These algorithms have decreasing efficiency as the subclass of systems gets bigger, with highest efficiency for underlying partial 2-tree graphs (alternately called tree-width 2, series-parallel, and $K_4$ minor avoiding), moderate efficiency for 1 degree-of-freedom (dof) graphs with low Cayley complexity (which include common linkages such as the Strandbeest, Limacon, and Cardioid), and decreased efficiency for general 1-dof tree-decomposable graphs. While software suites exist  \cite{keycurriculum1995geometer,porta2014open,siemens1999d,todd2007geometry}, no such formal algorithms and guarantees are known for non-tree-decomposable systems.

@@

\subsection{Using Convex Cayley Configuration Spaces}
\label{sec:2-tree-reduction}
%
Next we provide the necessary background to describe a specific approach for achieving the requirements (a)--(c) mentioned above, by restricting the class of reduced graphs $G' = G\setminus E_1$ and their isostatic completions $H$ in the above definition of the OMD problem, and using a key theorem of Convex Cayley configuration spaces \cite{sitharam2010convex}. This theorem characterizes the class of graphs $H$ and non-edges $F$ (Cayley parameters), such that the set of vectors $\lambda_F$ of  attainable lengths of the non-edges $F$ is always convex for any given lengths $\delta$ for the edges of $H$ (i.e.\ over all the realizations of the bar-joint constraint system or linkage $(H,\delta)$ in 2 dimensions). This set is called the (2-dimensional) \dfn{Cayley configuration space} of the linkage $(H,\delta)$ on the Cayley parameters $F$, denoted $\Phi_F(H,\delta)$ and can be viewed as a ``projection'' of the cartesian realization space of $(H,\delta)$ on the Cayley parameters $F$. Such graphs $H$ are said to have \dfn{convexifiable Cayley configuration spaces for some parameters $F$} (short: $H$ is \dfn{convexifiable} \todo{This shortening doesn't exactly make sense}).

@@

\begin{theorem*}\label{theorem:criterionc}
    [Efficient search]
    For an indecomposable, $k$-approximately convexifiable graph $G = (V,E)$, let $G' = (V,E' =E\setminus D)$ be a spanning partial 2-tree subgraph where $|D| \le  k$. Let  $F$ be a set of non-edges of $G$ such that $H = (V, E'\cup F)$ is a 2-tree. Each solution $p$ (or connected component of a solution space) of $(G,\delta)$ of an orientation type $\sigma_p$ can be found in time $O(\log(W))$ where $W$ is the number of cells of desired accuracy (discrete volume) of the convex polytope $\Phi_F(G',\delta_E')$. The (discrete) volume $W$ is exponential in $|F|$ and polynomial in the (discrete range) of the parameters in $F$.
\end{theorem*}

@@

    \caption{
    (\ref{fig:omd_k33_example:a}) The $K_{3,3}$ with two labeled edges, $e_1$ and $e_2$. (\ref{fig:omd_k33_example:b}) The $K_{3,3}$ with $e_1$ and $e_2$ removed (dashed lines) and rearranged to illustrate that it is now a partial 2-tree. (\ref{fig:omd_k33_example:c}) The $K_{3,3}$ with $\{e_1,e_2\}$ removed and $\{f_1,f_2\}$ (bold lines) added to make a 2-tree, showing that the $K_{3,3}$ is at least $OMD_2$. (\ref{fig:omd_k33_example:d}) The $K_{3,3}$ with only $e_2$ removed (dashed line). (\ref{fig:omd_k33_example:e}) The $K_{3,3}$ with $e_2$ removed and $f_3$ (bold line) added to make a low Cayley complexity graph, showing that the $K_{3,3}$ is $OMD_1$.
    }
    \label{fig:omd_k33_example}
\end{figure*}

@@

\myexample
[Using Cayley configuration space].
A graph $G=K_{3,3}$  cannot be decomposed into any nontrivial isostatic graphs, i.e.\ its DR-plan has a root and 9 children corresponding to the 9 edges. Solving or recombining the system $(G,\delta)$ corresponding to the root of this DR-plan involves solving a quadratic system with 8 equations and variables. Instead of simultaneously solving this system, we could instead use the fact that $G=K_{3,3}$ is in OMD$_2$: remove the edges $e_1,e_2$ in Figure \ref{fig:omd_k33_example} to give a partial 2-tree $G'$. Now add the non-edges $f_1,f_2$ to give a 2-tree $H$ with a DR-plan of size 2. The Cayley configuration space $\Phi_f(G', \delta_{E\setminus e})$ is a single interval of attainable length values $\lambda_F$ for the edge $f$. When $\delta$ is generic, i.e.\ does not admit collinearities or coincidences in the realizations of $(G,\delta)$, the realization space of $(H, \left<\delta_{E\setminus e}, \lambda_f\right>)$ has 16 solutions $q^p_{\lambda_f}$ (modulo orientation preserving isometries), with distinct orientation types $\sigma_p$ (two orientation choices for each of the 4 triangles) that can be obtained by solving a sequence of 4 single quadratics in 1 variable (DR-plan of size 2). By subdivided binary search in the interval $\lambda_f \in \Phi_f(G', \delta_{E\setminus e})$, the desired solution $p$  of $(G,\delta)$ is found when the length of the nonedge $e$ in  the realization $q^p_{\lambda_f}$ is $\delta_e$.

In fact, we can show that $G=K_{3,3}$ is in OMD$_1$ by removing a single edge to reduce (as shown in Figure \ref{fig:omd_k33_example}) to a tree-decomposable graph of low Cayley complexity (which includes the class of partial 2-trees). In the next Section \ref{sec:tdecomp}, we discuss this issue of why the largest class of reduced graphs is desirable.

\subsection{Optimized Modification by Enlarging the class of Reduced
Graphs}
\label{sec:tdecomp}
%
It is possible in principle to decrease $k$ for a OMD$_k$ graph (i.e.\ the number of edges to be removed to ensure an isostatic completion that is decomposable with a small DR-plan) by considering reduced graphs $G'$ (and modified graphs $H$) that come from a larger class than partial 2-trees but nevertheless have convex Cayley configuration spaces at least when the realization space is restricted to a sufficiently comprehensive orientation type. In particular, the so-called \dfn{tree-decomposable graphs of low Cayley complexity} \cite{sitharam2011cayleyI,sitharam2011cayleyII} include the partial 2-trees and many others that are not partial 2-trees. See an illustration in Figure \ref{fig:omd_k33_example}. These too result in DR-plans of size 2 or 3, putting $G$ in the class OMD$_k$ and thus meeting Criteria (a) and (b). The Criterion (c) is met---for example when $k=1$---because 1-dof Cayley configuration spaces of linkages based on such graphs $G'$ are known to be single intervals when a comprehensive orientation type $\sigma_p$ of the sought solution $p$ is given. In addition, the bounds of these intervals are of low algebraic complexity. More precisely, the bounds  can themselves be computed using a DR-plan of size 2 or 3, i.e.\ the computation of these bounds is tree-decomposable. Alternatively, the bounds are in a simple quadratic or radically solvable extension field over the rationals, or they can be computed by solving a triangularized system of quadratics.

>>>>

\subsubsection{Optimal Recombination and Solution Space Navigation}
For the entire DR-plan, finding all desired solutions is barely tractable even if recombination of solved subsystems is easy for each indecomposable parent system in the DR-plan. This is because even for the simplest, highly decomposable systems with small DR-plans, the problem of finding even a single solution to the input system at the root of the DR-plan is NP-hard  \cite{saxe1979embeddability} and there is a combinatorial explosion of solutions \cite{borcea2004number}. Typically, however, the desired solution has a given orientation type, in which case, the crux of the difficulty is concentrated in the algebraic complexity of (re)combining child system solutions to give a solution to the parent system at any given level of the DR-plan. For fairly general 3D constraint systems, there are algorithms with formal guarantees that uncover underlying matroids to combinatorially obtain an optimal parameterization to minimize the algebraic complexity of the indecomposable parent (sub)systems that occur in the DR-plan \cite{sitharam2010optimized,sitharam2006well,sitharam2010reconciling}, provided the DR-plan meets some of the abovementioned criteria.

@@

\subsubsection{Configuration Spaces of Underconstrained Systems}
For underconstrained 2D bar-joint and body-hyperpin qusecs obtained from various subclasses of tree-decomposable systems, algorithms have been developed to complete them into isostatic systems
\cite{joan-arinyo2003transforming,sitharam2005combinatorial,gao2006ctree,sitharam2010convex} and to find paths within the connected components \cite{sitharam2011cayleyI,hidalgo2011reachability} of standard Cartesian configuration spaces. Most of the algorithms with formal guarantees leverage Cayley configuration space theory \cite{sitharam2010convex,sitharam2011cayleyI,sitharam2011cayleyII} to characterize subclasses of graphs and additional constraints that control combinatorial explosion, and provide faithful bijective representation of connected components and paths. These algorithms have decreasing efficiency as the subclass of systems gets bigger, with highest efficiency for underlying partial 2-tree graphs (alternately called tree-width 2, series-parallel, and $K_4$ minor avoiding), moderate efficiency for 1 degree-of-freedom (dof) graphs with low Cayley complexity (which include common linkages such as the Strandbeest, Limacon, and Cardioid), and decreased efficiency for general 1-dof tree-decomposable graphs. While software suites exist  \cite{keycurriculum1995geometer,porta2014open,siemens1999d,todd2007geometry}, no such formal algorithms and guarantees are known for non-tree-decomposable systems.

@@

\subsection{Using Convex Cayley Configuration Spaces}
\label{sec:2-tree-reduction}
%
Next we provide the necessary background to describe a specific approach for achieving the requirements (a)--(c) mentioned above, by restricting the class of reduced graphs $G' = G\setminus E_1$ and their isostatic completions $H$ in the above definition of the OMD problem, and using a key theorem of Convex Cayley configuration spaces \cite{sitharam2010convex}. This theorem characterizes the class of graphs $H$ and non-edges $F$ (Cayley parameters), such that the set of vectors $\lambda_F$ of  attainable lengths of the non-edges $F$ is always convex for any given lengths $\delta$ for the edges of $H$ (i.e.\ over all the realizations of the bar-joint constraint system or linkage $(H,\delta)$ in 2 dimensions). This set is called the (2-dimensional) \dfn{Cayley configuration space} of the linkage $(H,\delta)$ on the Cayley parameters $F$, denoted $\Phi_F(H,\delta)$ and can be viewed as a ``projection'' of the cartesian realization space of $(H,\delta)$ on the Cayley parameters $F$. Such graphs $H$ are said to have \dfn{convexifiable Cayley configuration spaces for some parameters $F$} (short: $H$ is \dfn{convexifiable}).

@@

\begin{theorem*}[Efficient search]
\label{theorem:criterionc}
    % [Efficient search]
    For an indecomposable, $k$-approximately convexifiable graph $G = (V,E)$, let $G' = (V,E' =E\setminus D)$ be a spanning partial 2-tree subgraph where $|D| \le  k$. Let  $F$ be a set of non-edges of $G$ such that $H = (V, E'\cup F)$ is a 2-tree. Each solution $p$ (or connected component of a solution space) of $(G,\delta)$ of an orientation type $\sigma_p$ can be found in time $O(\log(W))$ where $W$ is the number of cells of desired accuracy (discrete volume) of the convex polytope $\Phi_F(G',\delta_E')$. The (discrete) volume $W$ is exponential in $|F|$ and polynomial in the (discrete range) of the parameters in $F$.
\end{theorem*}

@@

    \caption{
    (\ref{fig:omd_k33_example:a}) The $K_{3,3}$ with two labeled edges, $e_1$ and $e_2$. (\ref{fig:omd_k33_example:b}) The $K_{3,3}$ with $e_1$ and $e_2$ removed (dashed lines) and rearranged to illustrate that it is now a partial 2-tree. (\ref{fig:omd_k33_example:c}) The $K_{3,3}$ with $\{e_1,e_2\}$ removed and $\{f_1,f_2\}$ (bold lines) added to make a 2-tree, showing that the $K_{3,3}$ is at least OMD$_2$. (\ref{fig:omd_k33_example:d}) The $K_{3,3}$ with only $e_2$ removed (dashed line). (\ref{fig:omd_k33_example:e}) The $K_{3,3}$ with $e_2$ removed and $f_3$ (bold line) added to make a low Cayley complexity graph, showing that the $K_{3,3}$ is OMD$_1$.
    }
    \label{fig:omd_k33_example}
\end{figure*}

@@

\begin{example}[Using Cayley configuration space]
% \myexample
% [Using Cayley configuration space].
A graph $G=K_{3,3}$  cannot be decomposed into any nontrivial isostatic graphs, i.e.\ its DR-plan has a root and 9 children corresponding to the 9 edges. Solving or recombining the system $(G,\delta)$ corresponding to the root of this DR-plan involves solving a quadratic system with 8 equations and variables. Instead of simultaneously solving this system, we could instead use the fact that $G=K_{3,3}$ is in OMD$_2$: remove the edges $e_1,e_2$ in Figure \ref{fig:omd_k33_example} to give a partial 2-tree $G'$. Now add the non-edges $f_1,f_2$ to give a 2-tree $H$ with a DR-plan of size 2. The Cayley configuration space $\Phi_f(G', \delta_{E\setminus e})$ is a single interval of attainable length values $\lambda_F$ for the edge $f$. When $\delta$ is generic, i.e.\ does not admit collinearities or coincidences in the realizations of $(G,\delta)$, the realization space of $(H, \left<\delta_{E\setminus e}, \lambda_f\right>)$ has 16 solutions $q^p_{\lambda_f}$ (modulo orientation preserving isometries), with distinct orientation types $\sigma_p$ (two orientation choices for each of the 4 triangles) that can be obtained by solving a sequence of 4 single quadratics in 1 variable (DR-plan of size 2). By subdivided binary search in the interval $\lambda_f \in \Phi_f(G', \delta_{E\setminus e})$, the desired solution $p$  of $(G,\delta)$ is found when the length of the nonedge $e$ in  the realization $q^p_{\lambda_f}$ is $\delta_e$.

In fact, we can show that $G=K_{3,3}$ is in OMD$_1$ by removing a single edge to reduce (as shown in Figure \ref{fig:omd_k33_example}) to a tree-decomposable graph of low Cayley complexity (which includes the class of partial 2-trees). In the next Section \ref{sec:tdecomp}, we discuss this issue of why the largest class of reduced graphs is desirable.
\end{example}

\subsection{Optimized Modification by Enlarging the Class of Reduced
Graphs}
\label{sec:tdecomp}
%
It is possible in principle to decrease $k$ for a OMD$_k$ graph (i.e.\ the number of edges to be removed to ensure an isostatic completion that is decomposable with a small DR-plan) by considering reduced graphs $G'$ (and modified graphs $H$) that come from a larger class than partial 2-trees but nevertheless have convex Cayley configuration spaces at least when the realization space is restricted to a sufficiently comprehensive orientation type. In particular, the so-called \dfn{tree-decomposable graphs of low Cayley complexity} \cite{sitharam2011cayleyI,sitharam2011cayleyII} include the partial 2-trees and many others that are not partial 2-trees. See an illustration in Figure \ref{fig:omd_k33_example}. These too result in DR-plans of size 2 or 3, putting $G$ in the class OMD$_k$ and thus meeting Criteria (a) and (b). The Criterion (c) is met---for example when $k=1$---because 1-dof Cayley configuration spaces of linkages based on such graphs $G'$ are known to be single intervals when a comprehensive orientation type $\sigma_p$ of the sought solution $p$ is given. In addition, the bounds of these intervals are of low algebraic complexity. More precisely, the bounds  can themselves be computed using a DR-plan of size 2 or 3, i.e.\ the computation of these bounds is tree-decomposable. Alternatively, the bounds are in a simple quadratic or radically solvable extension field over the rationals, or they can be computed by solving a triangularized system of quadratics.

### 2015-05-17 doc/GMP2015/revision_2/s05_problems.tex
Revision 2, finished.

\subsubsection{Special classes of small DR-plans}
As seen in the previous section, 2-trees and tree-decomposable graphs have not only small, but also special DR-plans that permit easy solving---essentially by solving a single quadratic at a time.

@@

%In the recombination setting, the corresponding \dfn{restricted
%OMD$_k$  problem} requires the reduced graph $G'$ and its isostatic
%completion $H$ to be 2-trees as in Section \ref{sec:2-tree-reduction}
%or to be a low Cayley complexity tree-decomposable graph as in Section
%\ref{sec:tdecomp}. Clearly these problems have deterministic
%polynomial time algorithms, the algorithms run in time exponential in
%$k$. As a result, when $k$ is part of the input, the \dfn{restricted
%OMD problem} has the potential to be difficult. For example, when the
%isostatic completion $H$ is required to be a 2-tree the restricted OMD
%problem is reducible to the maximum spanning series-parallel subgraph
%problem shown by \cite{cai1993spanning} to be NP-complete even if the input
%graph is planar of maximum degree at most 6. However, since the OMD
%problem has other input restrictions such as not having any proper
%isostatic subgraphs, it is not clear if the reverse reduction exists
%and hence it is unclear whether the OMD problem is NP-complete. The
%same holds for the restricted OMD problem where the isostatic
%completion $H$ is required to be a tree-decomposable graph of low
%Cayley complexity (i.e, have special, small DR-plans). One potential
%obstacle to an indecomposable graph $G$'s membership in the restricted
%OMD$_k$ for small $k$ is if $G$ is tri-connected and has large girth.
%In fact, 6-connected (hence rigid) graphs with arbitrarily large girth
%have been constructed in \cite{servatius2000rigidity}.
%
\subsubsection{Optimal Modification, Completion and Recombination: previous work and formal
connections}
%
The OMD problem is closely related to a well-studied problem of completion of an underconstrained system to an isostatic one with a small DR-plan.
\begin{observation}\label{obs:OC_to_OMD}
    The (decision version of) the \textbf{optimal completion problem (OC)} from \cite{sitharam2005combinatorial,joan-arinyo2003transforming,zhang-gao2006well} is OMD$_0$.
\end{observation}
In fact, a \dfn{restricted OC} problem was studied by \cite{joan-arinyo2003transforming} requiring the completion to be tree-decomposable.
%It is not clear if
%the reverse reduction holds, i.e, is OMD reducible to OC?

>>>>

\subsubsection{Special Classes of Small DR-Plans}
As seen in the previous section, 2-trees and tree-decomposable graphs have not only small, but also special DR-plans that permit easy solving---essentially by solving a single quadratic at a time.

@@

%In the recombination setting, the corresponding \dfn{restricted
%OMD$_k$  problem} requires the reduced graph $G'$ and its isostatic
%completion $H$ to be 2-trees as in Section \ref{sec:2-tree-reduction}
%or to be a low Cayley complexity tree-decomposable graph as in Section
%\ref{sec:tdecomp}. Clearly these problems have deterministic
%polynomial time algorithms, the algorithms run in time exponential in
%$k$. As a result, when $k$ is part of the input, the \dfn{restricted
%OMD problem} has the potential to be difficult. For example, when the
%isostatic completion $H$ is required to be a 2-tree the restricted OMD
%problem is reducible to the maximum spanning series-parallel subgraph
%problem shown by \cite{cai1993spanning} to be NP-complete even if the input
%graph is planar of maximum degree at most 6. However, since the OMD
%problem has other input restrictions such as not having any proper
%isostatic subgraphs, it is not clear if the reverse reduction exists
%and hence it is unclear whether the OMD problem is NP-complete. The
%same holds for the restricted OMD problem where the isostatic
%completion $H$ is required to be a tree-decomposable graph of low
%Cayley complexity (i.e, have special, small DR-plans). One potential
%obstacle to an indecomposable graph $G$'s membership in the restricted
%OMD$_k$ for small $k$ is if $G$ is tri-connected and has large girth.
%In fact, 6-connected (hence rigid) graphs with arbitrarily large girth
%have been constructed in \cite{servatius2000rigidity}.
%
\subsubsection{Optimal Modification, Completion and Recombination: Previous Work and Formal Connections}
%
The OMD problem is closely related to a well-studied problem of completion of an underconstrained system to an isostatic one with a small DR-plan.
\begin{observation}\label{obs:OC_to_OMD}
    The (decision version of) the \textbf{optimal completion problem (OC)} from \cite{sitharam2005combinatorial,joan-arinyo2003transforming,zhang-gao2006well} is OMD$_0$.
\end{observation}
In fact, a \dfn{restricted OC} problem was studied by \cite{joan-arinyo2003transforming} requiring the completion to be tree-decomposable.
%It is not clear if
%the reverse reduction holds, i.e, is OMD reducible to OC?

### 2015-05-17 doc/GMP2015/revision_2/s07_pinnedline.tex
Revision 2, finished.

\section{Application: Finding optimal DR-plans and realizations for cross-linking microfibrils }
\label{sec:pinnedline}

@@

\subsection{Modeling the fibrils as a pinned line-incidence system}

@@

\subsection{Optimal DR-plan for pinned line-incidence systems}

>>>>

\section{Application: Finding Optimal DR-Plans and Realizations for Cross-Linking Microfibrils }
\label{sec:pinnedline}

@@

\subsection{Modeling the Fibrils as a Pinned Line-Incidence System}

@@

\subsection{Optimal DR-Plan for Pinned Line-Incidence Systems}
